A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation

### 2018-12-02 Papers/AppliedMaths2018/Resubmission/Stochastic_Upscaling_UK_AMM_2018.tex
Updated manuscript 1.5 of 7 completed

\usepackage{amsthm}
\usepackage{graphicx}
%\usepackage{subfigure}

@@

\title{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
Singular Value Decomposition (SVD)\sep Upscaling\sep Model Order Reduction (MOR) \sep  Principal Component Spaces \sep Permeability field \sep Heterogeneity
\end{keyword}

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed by SVD in principal component spaces.

\bigskip
This work focuses on reducing the order of permeability distribution in porous media flow simulations. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a novel high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017} which couples both the fine resolution and upscaled representation of the permeability field. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used.

Four upscaling techniques are investigated in this work, arithmetic and harmonic means are used to obtain the first two upscaled representation and these two are classed as deterministic techniques. The third is a randomly generated permeability field with Gaussian distribution, prescribed by a probability density function (PDF) which was obtained from a domain discretised with high-resolution mesh with known permeability distribution (\ie base case). The fourth upscaling technique is the crux of this research, it introduces singular value decomposition (SVD) as a method of upscaling using the concept of principal component space to reduce the order of the permeability distribution.

In Section~\ref{section:svd}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:svd} describes model order space reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} starts by describing the high resolution (\ie base case) on which the four upscaling techniques are applied to. A brief summary of the pre-processing step for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion on the results of the simulations, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, Conclusions are drawn in Section~\ref{section:conclusion}.

In this work, a novel control volume finite element method (CVFEM) formulation is used to discretise and solve the set of multi-fluid flow equations. Continuity equations are embedded into the pressure equation to enforce mass conservation and the exact force balance (extended Darcy equations). The numerical formulation employs an implicit algorithm with respect to time which is less restrictive than the traditional industry-standard implicit-pressure-explicit-saturation (IMPES) scheme \cite{aziz_1986,geiger_2004}.

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{salinas2015,adam_2016}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{jackson_2013,Gomes_2017}).

\section{Brief Overview of Upscaling}\label{section:overview_upscaling}

Section~\ref{section:intro} briefly introduced the rational behind upscaling. Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This subject of upscaling heterogenous properties has interested industries (\ie oil and gas, CCS and waste management) as well as the academic community for over $7$ decades.

@@

\section{Singular Value Decomposition (SVD) and Application for Reducing the Order of Permeability Fields}\label{section:svd}
\subsection{A Brief Introduction to SVD}\label{subsection:svd_brief}
Given a matrix $\mathbf{A}\in\mathcal{R}^{m\times n}$, the singular value decomposition (SVD) is a method for factorizing $\mathbf{A}$, such that,
\begin{equation}
  \mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^{\intercal}.
\end{equation}
where $\mathbf{U}=\left[\underline{u}_{1},\underline{u}_{2},\cdots,\underline{u}_{m}\right]\in\mathcal{R}^{m\times n}$ and $\mathbf{V}=\left[\underline{v}_{1},\underline{v}_{2},\cdots,\underline{v}_{n}\right]\in\mathcal{R}^{m\times n}$ are orthogonal but not necessarily the same. $\Sigma$ is a diagonal matrix, \ie
\begin{displaymath}
  \Sigma=\text{diag}\left(\sigma_{1},\sigma_{2},\cdots,\sigma_{r}\right)\in\mathcal{R}^{m\times n},
\end{displaymath}
and $r=\min\left\{m,n\right\}$ is the rank of $\mathbf{A}$, with $\sigma_{1}\ge\sigma_{2}\ge\cdots\sigma_{r}\ge 0$. The diagonal entries of $\Sigma$ $\left(\text{\ie} \sigma_{i}, i=1,r\right)$ are called singular values of $\mathbf{A}$. If $\mathbf{A}$ is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are also square matrices. Also, as $\mathbf{U}$ and $\mathbf{V}$ are orthogonals and unitary, 
\begin{equation}
  \mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I},
\end{equation}
where $\mathbf{I}$ is an identity matrix. However, if $\mathbf{A}$ is an $m\times n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m\times m$ while $\mathbf{V}$ is an $n\times n$. Another relevant property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether $\mathbf{A}$ is square or rectangular) is that
\begin{equation}
  \mathbf{U}^{-1} = \mathbf{U}^{\intercal} \;\;\;\text{ and }\;\;\; \mathbf{V}^{-1} = \mathbf{V}^{\intercal},
\end{equation}
and if $\mathbf{A}$ is an $m \times n$ matrix thus $\Sigma$ is a diagonal $m \times n$ matrix.

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
In 3D geometry, permeability is a tensor quantity, but for the purpose of the following explanation (and the remaining of this paper), a 2D geometry is assumed where permeability is a scalar quantity. Furthermore, the simulator prescribes the permeability field by assigning permeability values to the centre of element-pairs \cite{Christou_2018} which discretises the 2D simulation domain.

Permeability values assigned in the centre of the element/cell represents values of permeability for the associated element $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this section, it will be assumed that the $2D$ geometry is discritized by a square grid with the value for each square element assigned in the centre of the element. This explanation allow one to easily visualize this permeability space because it is defined by the cartesian coordinates system.

The model order reduction for the permeability field discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data points are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through SVD in the principal component spaces.

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed through SVD to principal component spaces (PCS) where traditional principal component analysis is performed before interpolation is used to implement further data reduction in the principal component spaces. This reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space.

The PCS is more challenging to visualize because it usually consists of more than 3-D. An aid for visualizing the PCS is to write out the $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{U}$) as separate PCS.

\subsubsection{Model Order Reduction/Principal Component Space Reduction}\label{subsubsection:svdcase_preprocess_algorithm}
For the SVD model order reduction of the permeability field, all the work is done in the principal component spaces. This can be completed in two steps: the first step is the principal component analysis (PCA) \cite{Hotelling_1933}, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$.

In the case with $z = r$, it can be assumed that step 1 has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .

The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) to reduce the data points in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

@@

In both sets of test-cases, averages are calculated for each block using the $1600$ data points from the blocks of the BaseCase to obtain a permeability value for each block in the upscaled models. This effectively homogenizes the permeability field in each block. The upscaled permeability field for the third and fourth cases are referred to as PDFCase and SVDCase, respectively. And, as can be deduced from the names, these cases are based on probability density function (PDF) and SVD, respectively. The description for these permeability field models are presented in Sections~\ref{subsubsection:pdfcase_preprocess_algorithm} and ~\ref{subsubsection:svdcase_preprocess_algorithm} respectively. It is appropriate to note that although these models (PDFCase and SVDCase) are upscaled, they retain heterogeneity within the blocks.

@@

\subsubsection{SVDCase Pre-processing Algorithm}\label{subsubsection:svdcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item The permeability field obtained in step 3 of the pre-processing algorithm for the base case is factorized using SVD to obtain $\mathbf{U}, \Sigma,$ and $\mathbf{V}$.
  \item A selected number ($z$) of singular values are retained in decreasing order of magnitude to form a new diagonal matrix, $\Sigma_{new}$.
  \item The first $z$ columns of $\mathbf{U}$ and $\mathbf{V}$ are selected to form new truncated matrices call $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$ respectively.
  \item Within each column of $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$, a linear interpolation is performed to form a new column with the number of data points in each column reduced to half of the original column. These new columns are recombined to form $\mathbf{U_{new2}}$ and $\mathbf{V_{new2}}$.
  \item The permeability field used for the SVD case is the matrix $\mathbf{A_{new2}}$, which is defined as the product of $\mathbf{U_{new2}} \mathbf{\Sigma_{new}} \mathbf{V_{new2}^{\intercal}}$. It is a permeability field with a quarter of the data values in the base case and therefore a permeability field with reduced order.  
\end{enumerate}

@@

This highlights the fact that the permeability distribution of the ArithMean and HarmMean cases, Fig.~\ref{fig:PermFields}(c) and (d), are homogeneous within each block, while PDFCase and SVDCase cases, Fig.~\ref{fig:PermFields}(e) and (f), are heterogeneous. Given that the BaseCase is also heterogeneous, it can be stated that the model for the PDFCase and the SVDCase give a better representation of the BaseCase. An additional consequence of the homogeneity in the ArithMeanCase and the HarmMeanCase is that the upgrid size has no influence on the outcome of their permeability distribution. Whereas, for the PDFCase and the SVDCase, the permeability field will change if the upgrid resolution changes.

With respect to the PDFCase and for this specific resolution, it is important to note that Fig.~\ref{fig:PermFields}(e) is only one realiztion of infinitely many possible representations of the PDFCase. Meanwhile, the SVDCase will retain this specific (\ie Fig.~\ref{fig:PermFields}(e)) permeability field for this specific resolution (\ie if linear interpolation is used, but this digresses from the main point and will be discussed in future work). Therefore, the permeability value at a specific point in the PDFCase is not directly related to the data value at the nearest point in the BaseCase because the randomly generated PDF dataset is also randomly arranged within each block. However, the data value at a point in the SVDCase is directly related to the nearest data values with respect to the same point in the BaseCase, throught the interpolation which takes place in the PCS.

Furthermore, prior to the pre-processing required for evaluating the ArithMeanCase, HarmMeanCase and the PDFCase it is necessary to have an understanding of the whole domain and pre-define blocks with similar properties (\ie blocks $K1,~K2,~K3$ and $K4$) in this case. However, with the SVDCase, knowledge of blocks with similar properties is not required prior to the upscaling. Albeit, one has to select the upgrid resolution carefully, if the resolution is too large then there is the risk of losing relevant data in the interpolation. The limits of the techniques used for the SVDCase will be discussed in the author's future work.

@@

Figure~\ref{fig:Saturationfield@t=1.15s}(d) also shows an inconsistency with the BaseCase (Fig.~\ref{fig:Saturationfield@t=1.15s}(a)) where it shows phase 1 moving from block $K2$ into $K4$. This inconsistency is with respect to this realization of the PDFCase. Hence, SVDCase in Fig.~\ref{fig:Saturationfield@t=1.15s}(e) give the best representation of the BaseCase in \ie Fig.~\ref{fig:Saturationfield@t=1.15s}(a)

@@

Upscaling is a form of model order reduction, and a method of implementing upscaling using singular value decomposition, SVD, is introduced in this work. The SVD method of upscaling first factorizes the matrix of the permeability field and then performs interpolation within the principal component space. The interpolated data in the principal component space is then recombined to form a reduced matrix of the permeability field.

Results show that the permeability field data reduction can be specifically defined for the PDFCase and the SVDCase but not for the ArithMeanCase and the HarmMeanCase. This is because the latter cases will produce the same permeabiBuckleylity field irrespective of the upgrid resolution. The SVDCase is the only technique that does not require a prior knowledge of blocks with similar permeability values before upscaling.

The author's future work will provide a statistical analysis of the technique used for the SVDcase and also presents other implementations of the technique.

>>>>

\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[mathscr]{euscript}
%\usepackage{subfigure}

@@

\title{A Reduced Order Model for Porous Media Flows which Couples Linear Regression with Principal Component Analysis (PCA) using SVD or HOSVD: Upscaling the Permeability Field for Multiphase Reservoir Flow Simulation}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

In this paper, a single realization of a permeability field was upscaled by first projecting the permeability field into a principal component analysis space (using SVD for 2D field and HOSVD for 3D fields). Then linear regression was employed within each of the principal component spaces to obtain approximate representations of the initial principal component. The new principal components are then recombined to for a new permeability field with a reduced order. This PCA-linear regression method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance.

It was shown that the new upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
Principal Component Analysis (PCA) and Linear Regression \sep Singular Value Decomposition (SVD) and Higer Order Singular Value Decomposition (HOSVD) \sep Upscaling\sep Reduced Order Models (ROM), \sep Permeability field \sep Heterogeneity
\end{keyword}

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis (PCA) \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition (POD) \cite{Sirovich_1987} are also MOR methods discussed in the references published in 1984 and 1987, respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed through linear regression in the PCA space which is obtained with SVD or HOSVD for 2D and 3D cases, respectively.

\bigskip
This work focuses on reducing the order of permeability distribution in porous media flow simulations. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a novel high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017} which couples both the fine resolution and upscaled representation of the permeability field. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used. Four upscaling techniques are investigated in this work and they are named Arithmetric mean, Harmonic mean, probability density function (PDF) and PCA (using SVD for 2D models and HOSVD for 3D model) coupled with linear regression.

In Section~\ref{section:pca_linreg}, fundamental mathematics of SVD and associated properties are presented. Then, the concept of spaces is explained to give the reader the understanding necessary to appreciate the transformations and projections involved in this work. Finally, Section~\ref{section:pca_linreg} describes the reduced order model reduction for the permeability field within the principal component space. Section~\ref{section:model_simulation} is dedicated to model description and simulation. It starts by describing the high resolution (\ie base case) on which the four upscaling techniques are applied to. A brief summary of the pre-processing step for each of the models is also presented. Section~\ref{section:results_discussion} provides the results and further discussion on the results of the simulations, it also gives further interpretation to the result obtained from the SVD upscaling technique. Finally, Conclusions are drawn in Section~\ref{section:conclusion}.

\subsection{Brief Overview of Upscaling}\label{section:overview_upscaling}

Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This subject of upscaling heterogenous properties has interested industries (\ie oil and gas, CCS and waste management) as well as the academic community for over $7$ decades.

@@

In this work, a control volume finite element method (CVFEM) formulation is used to discretise and solve the set of multi-fluid flow equations. Continuity equations are embedded into the pressure equation to enforce mass conservation and the exact force balance (extended Darcy equations). The numerical formulation employs an implicit algorithm with respect to time which is less restrictive than the traditional industry-standard implicit-pressure-explicit-saturation (IMPES) scheme \cite{aziz_1986,geiger_2004}.

\medskip
The discretised saturation and Darcy equations are solved using a multigrid-like approach as described by \citeauthor{pavlidis2016}~\cite{pavlidis2016}. The numerical formulation is fully described by \citeauthor{Gomes_2017}~\cite{Gomes_2017} (see also \cite{salinas2015,adam_2016}). These numerical methods are embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} (a full description of the model can be found in \cite{jackson_2013,Gomes_2017}). Although, the numerical formulation used here is relatively new, it is not the focus of this work. The result and discussion section (Section \ref{section:results_discussion}) refers to \citet{dawe_2008} for validation of the numerical formulation with respect to the 2D simulation. However, more extensive validation tests can be found in the aforementioned references \cite{adam_2016, Gomes_2017,jackson_2013, pavlidis2016, salinas2015}.

\section{PCA coupled with Linear Regression}\label{section:pca_linreg}

@@

\subsubsection{Understanding the permeability space}\label{subsubsection:visualization_permspace}
Permeability can be represented as tensor quantity, but for the purpose of the following explanation (and the remaining of this paper), it is assumed that permeability is a scalar quantity. Furthermore, the simulator prescribes the permeability field by assigning permeability values to the centre of element-pairs \cite{Christou_2018} which discretises the simulation domain.

Permeability values assigned in the centre of the element/cell represents values of permeability for the associated element $\left(\text{\ie P}_{0}\text{DG element-pair}\right)$. For simplicity, in the remaining part of this sub-section, a $2D$ geometry is assumed where the geometry is discritized by a square grid with the value for each square element assigned in the centre of the element. This explanation allow one to easily visualize this permeability space because it is defined by the cartesian coordinates system.

The ROM method discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data points are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through linear regression in the principal component spaces.

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed with SVD for 2D permeability field and HOSVD for 3D permeability fields, to principal component spaces (PCS). Within the PCS, traditional PCA dimensionality reduction can be optionally performed (\ie{feature selection or elimination}, see Section~\ref{subsubsection:pca_dimred_linreg} for more details), before interpolation is used to implement further data reduction in the principal component spaces. The reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space. An algorithm which impliments the procedure described here, is presented in Section~\ref{subsubsection:svdcase_preprocess_algorithm}

The PCS is a bit challenging to visualize because it represents projections which are embedded in the SVD/HOSVD factors of the original permeability field. An aid for visualizing the PCS for a 2D permeability field is to write out the $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or SVD factors) and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{V^{\intercal}}$) as separate PCS. A similar intuition can be extended to HOSVD for 3D permeability fields. The decomposition of a matrix into its factors using SVD  \cite{Cline_2007, Tharwat_2016} or HOSVD \cite{Kolda_2009, Lathauwer_2000} is covered in details in several published materials. However, for completeness, the appendix contains a brief introduction to SVD (\ref{subsection:svd_brief}) and HOSVD (\ref{subsection:hosvd_brief}).

\subsubsection{Reduced Order Model/Principal Component Space Reduction}\label{subsubsection:pca_dimred_linreg}
For the PCA-Linear regression (PCA-LinReg) reduced order model of the permeability field, all the work is done in the principal component spaces. This can be completed in two steps, which is exemplified for the 2D model thus:
\begin{enumerate}
  \item The first step is the principal component analysis (PCA) \cite{Hotelling_1933, Tharwat_2016} using the SVD method, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$.

Note that, in the case with $z = r$, it can be assumed that step 1 has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .

\item The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) to reduce the data points in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

\end{enumerate}

It is trivial to show that these two steps can be replicated for a 3-D model using HOSVD to obtain the PCA space.\\

Finally on the PCS, it is worth noting that two methods exist for obtaining PCA space\cite{Tharwat_2016} , namely, the co-variance method and the SVD method. \citet{Tharwat_2016} goes on to describe both methods in details. In this work, the authors have opted to use the SVD method to obtain the PCA space. This is because the singular vectors which make up the PCS have a strong correlation to the coordinate axis they represent. This was exploited in making the assumption that a linear relationship exist for the data in each of the PC spaces, hence the linear regression performed to reduce the data in the PCS.

\subsection{Principal Component Analysis Versus Linear Regression}\label{subsection:pcs_vs_linreg}

\citet{AndrewNg_2018} was quoted to have said ``PCA is not linear regression, and despite some cosmetic similarity, these are actually totally different algorithms''. The main differences between these two mathematical methods is summarised in the Table~\ref{table:pca_vs_linreg} below and can also be deduced from the figure below.

\begin{table}[h!]
\centering
\begin{tabular}{|m{1em} |m{14em} |m{14em}|} 
 \hline
  & PCA & Linear Regression \\ [0.5ex] 
 \hline
 1 & Determines the best vector (or set of vectors) which compresses the data points by minimizing the orthogonal distance to the points & Approximates the best fit line for the points by minimizing the error \\
 \hline
 2 & Essentially, it forms a projection of the original points onto the new vector space & Predicts $y$ from $x$ through a function $f(x_1,x_2,....,x_n) = y$ \\
 \hline
 3 & Not used for predictions, as original feature space (\ie $x_1$ and $x_2$) give way to a new reduced dimension. Cordinates $x_1, x_2,...,x_n$ give way to $z_1,....,z_k$ where $k \leq n$ & \\[1ex]
 \hline
\end{tabular}
\caption{Table comparing PCA with Linear Regression}
\label{table:pca_vs_linreg}
\end{table}

\red{I need to plot a figure here.}

@@

As previosly mentioned, four upscaling techniques are investigated in this work. Arithmetic and harmonic means are used to obtain the first two upscaled representation and these two are classed as deterministic techniques. The third is a randomly generated permeability field with Gaussian distribution, prescribed by a probability density function (PDF) which was obtained from a domain discretised with high-resolution mesh with known permeability distribution (\ie base case). The fourth upscaling technique is the crux of this research, it introduces a reduced order model which couples PCA and linear regression.

@@

In both sets of test-cases, averages are calculated for each block using the $1600$ data points from the blocks of the BaseCase to obtain a permeability value for each block in the upscaled models. This effectively homogenizes the permeability field in each block. The upscaled permeability field for the third and fourth cases are referred to as PDFCase and PCA-LinReg Case, respectively. And, as can be deduced from the names, these cases are based on probability density function (PDF) and SVD, respectively. The description for these permeability field models are presented in Sections~\ref{subsubsection:pdfcase_preprocess_algorithm} and ~\ref{subsubsection:svdcase_preprocess_algorithm} respectively. It is appropriate to note that although these models (PDFCase and PCA-LinReg Case) are upscaled, they retain heterogeneity within the blocks.

@@

\subsubsection{2-D PCA (\ie{using SVD}) \& LinReg Case Pre-processing Algorithm}\label{subsubsection:svdcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item The permeability field obtained in step 3 of the pre-processing algorithm for the base case is factorized using SVD to obtain $\mathbf{U}, \Sigma,$ and $\mathbf{V}$.
  \item A selected number ($z$) of singular values are retained in decreasing order of magnitude to form a new diagonal matrix, $\Sigma_{new}$. (This step is optional, it is the conventional PCA dimensionality reduction)
  \item The first $z$ columns of $\mathbf{U}$ and $\mathbf{V}$ are selected to form new truncated matrices call $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$ respectively. (This step is necessary only if the optional step in 2 above is executed)
  \item Within each column of $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$, a linear interpolation is performed to form a new column with the number of data points in each column reduced to half of the original column. These new columns are recombined to form $\mathbf{U_{new2}}$ and $\mathbf{V_{new2}}$.
  \item The permeability field used for the SVD case is the matrix $\mathbf{A_{new2}}$, which is defined as the product of $\mathbf{U_{new2}} \mathbf{\Sigma_{new}} \mathbf{V_{new2}^{\intercal}}$. It is a permeability field with a quarter of the data values in the base case and therefore a permeability field with reduced order.  
\end{enumerate}

\subsubsection{3-D PCA (\ie{using HOSVD}) \& LinReg Case Pre-processing Algorithm}\label{subsubsection:hosvdcase_preprocess_algorithm}
The algorithm for the 3-D case is similar to that of the 2-D case described above. However:
\begin{enumerate}[1.]
\item In step 1 above, HOSVD is used for factorization instead of SVD. Therefore, a $3 x 3$ core tensor is obtained in place of the $\mathbf{\Sigma}$ term, while 3 factor matrices are obtained in place of $\mathbf{U}$ and $\mathbf{V}$.
  \item It goes without saying that the core tensor is treated as the $\mathbf{\Sigma}$ term, while the 3 factor matrices are treated similarly to $\mathbf{U}$ and $\mathbf{V}$ as described in Section \ref{subsubsection:svdcase_preprocess_algorithm} above.
\end{enumerate}
The authors found the updated Tensorly library package \cite{Tensorly_2018} helpful in implimenting and executing the algorithm for HOSVD.

@@

This highlights the fact that the permeability distribution of the ArithMean and HarmMean cases, Fig.~\ref{fig:PermFields}(c) and (d), are homogeneous within each block, while PDFCase and PCA-LinReg cases, Fig.~\ref{fig:PermFields}(e) and (f), are heterogeneous. Given that the BaseCase is also heterogeneous, it can be stated that the model for the PDFCase and the PCA-LinReg Case give a better representation of the BaseCase. An additional consequence of the homogeneity in the ArithMeanCase and the HarmMeanCase is that the upgrid size has no influence on the outcome of their permeability distribution. Whereas, for the PDFCase and the PCA-LinReg Case, the permeability field will change if the upgrid resolution changes.

With respect to the PDFCase and for this specific resolution, it is important to note that Fig.~\ref{fig:PermFields}(e) is only one realiztion of infinitely many possible representations of the PDFCase. Meanwhile, the PCA-LinReg Case will retain this specific (\ie Fig.~\ref{fig:PermFields}(e)) permeability field for this specific resolution (\ie if linear interpolation is used, but this digresses from the main point and will be discussed in future work). Therefore, the permeability value at a specific point in the PDFCase is not directly related to the data value at the nearest point in the BaseCase because the randomly generated PDF dataset is also randomly arranged within each block. However, the data value at a point in the PCA-LinReg Case is directly related to the nearest data values with respect to the same point in the BaseCase, throught the interpolation which takes place in the PCS.

Furthermore, prior to the pre-processing required for evaluating the ArithMeanCase, HarmMeanCase and the PDFCase it is necessary to have an understanding of the whole domain and pre-define blocks with similar properties (\ie blocks $K1,~K2,~K3$ and $K4$) in this case. However, with the PCA-LinReg Case, knowledge of blocks with similar properties is not required prior to the upscaling. Albeit, one has to select the upgrid resolution carefully, if the resolution is too large then there is the risk of losing relevant data in the interpolation. The limits of the techniques used for the PCA-LinReg Case will be discussed in the author's future work.

@@

Figure~\ref{fig:Saturationfield@t=1.15s}(d) also shows an inconsistency with the BaseCase (Fig.~\ref{fig:Saturationfield@t=1.15s}(a)) where it shows phase 1 moving from block $K2$ into $K4$. This inconsistency is with respect to this realization of the PDFCase. Hence, PCA-LinReg Case in Fig.~\ref{fig:Saturationfield@t=1.15s}(e) give the best representation of the BaseCase in \ie Fig.~\ref{fig:Saturationfield@t=1.15s}(a)

@@

Upscaling is a form of reduced order model, and a method of implementing upscaling by coupling linear regression with PCA is introduced in this work. For a 2-D model, SVD is used to decompose the existing permeability field to obtain the PCA space, while HOSVD is used for 3-D models. Within, the PCA space, linear regression (or interpolation) is used to reduce the data in the PCA space. The reduced data in the principal component space is then recombined to form a reduced data of the permeability field.

Results show that the permeability field data reduction can be specifically defined for the PDFCase and the PCA-LinReg Case but not for the ArithMeanCase and the HarmMeanCase. This is because the latter cases will produce the same permeability field irrespective of the upgrid resolution. The PCA-LinReg Case is the only technique that does not require a prior knowledge of blocks with similar permeability values before upscaling. This is a huge advantage as it saves time during the data pre-processing stage, because sorting and calculations for each distinct region is avoided.

The author's future work will provide a quantitative analysis of the results obtained from the coupled PCA-Linear regression method. Additionally, future work will present other methods which follow a similar principle but also seek to optimize the quantitative results obtained.

%\section{Acknowledgements}
%Mr Babatunde Lashore
